Vibrations and Heat Diffusion on the Unit Disc

### 2020-04-30 writeup/draft2.tex
fixed typos and added appendix

\documentclass{article}
\usepackage{changepage}
\usepackage{float}
\usepackage{noah}

@@

    F_{net}=T \sin(\theta + d\theta)-T \sin(\theta)=T\cos(\theta)d\theta)

@@

where $T$ is the tension in the rope. If we assume small enough displacements, we can approximated $\cos\theta\approx 1$ and write $d\theta= (\partial\theta/\partial x) dx$. Since we are assuming small displacements, we can say that the angle $\theta$ is the slope of the string. We will write the displacement of the string as $u=u(x,t)$, and say that this function must satisfy the following equation:

@@

Here we have absorbed the thermal conductivity into our definition of time. The interesting part of this equation is that it can be used to describe diffusive processes as well. In fact it can be used to describe many systems that do no violate the first and second law of thermodynamics. The first of these states energy conservation, and the second states that the entropy of a system must increase. This second law is one of the few that does depend on the direction of time. This is why the heat equation is not symmetric in the spatial and temporal variables, when the wave equation is.\\

@@

It is important to notice that for both of these equations, the spacial part is the same. Given that we are using the same boundary conditions for both equations, solving this equation will provide us with solutions to both the wave and heat equation.

@@

    \left(f(r,\phi),g(r,\phi)\right) = \int_0 ^1 dr \int_0 ^{2\pi} dr f^\star(r,\phi) g(r,\phi) r

@@

Before we continue, it is important to note that the normal modes that oscillate more spatially, (those with a larger eigenvalue) will damp more quickly than others. This shows the smoothing property of solutions to the heat equation. Given that our boundary data states that the function must vanish at the boundary, we see that the steady state solution is one with uniform temperature. If the boundary was non-uniform, then we would find that the solution would evolve to a soution of Laplace's equation whose boundary values matched the boundary temperature.\\

@@

In a similar vain, we may write the solution to the wave equation in the following way:

@@

\begin{figure}[H]
  \centering
  \includegraphics[width = 1.0\textwidth]{images/wavevel.png}
  \caption{Evolution of wave equation for initial velocity given.}
\end{figure}
We can see that these solutions to the heat equation are oscillating about the steady state solution $u = 0$. We also see that for the solution where only the initial velocity is given, the amplitude is much lower than it is when the initial displacement is given. This is because we are dividing each term of the solution by its corresponding eigenvalue. This in turn means that the initial velocity does not impact the dynamics as much as the initial position does.
\section{Conclusion}
Through the method of eigenfunction expansions, solutions to the heat and wave equation are generated, which in turn allow us to analyze the dynamics of solutions. We see that the wave equation is the higher dimensional analogue to the simple harmonic oscillator, and the heat equation is the higher dimensional analogue to exponential decay. Instead of the solutions oscillating or decay towards and equilibrium point, they end up decaying to an equilibrium function, one that is harmonic.\\
It is important to reiterate that this method of solving functions by eigenfunction expansion is not unique to Laplace operator, and can be used for any operator that satisfies equation (\ref{selfadjoint}), known as a self-adjoint operator. This allows us to in principal solve nearly any linear-homogenous PDE. In fact this method has a direct analogy to methods for solving a linear system of ODEs. Traditionally, we attempt to find the eigenvalues and straight-line solutions to the equation, which corresponds to finding the normal modes of a differential operator.
\newpage
\section{References}
Asmar, N. H. (2002). \textit{Applied complex analysis with partial differential equations}. New Jersey: Prentice-Hall Inc.\\
\begin{adjustwidth}{2 cm}{}
This book was used to understand differential equations in polar and spherical coordinate systems, as well as certain properties of Bessel functions.\\
\end{adjustwidth}
Farlow, S. J. (2016). \textit{Partial differential equations for scientists and engineers}. New York: Dover Publications, Inc.\\
\begin{adjustwidth}{2 cm}{}
This book was used to understand the basic types of partial differential equations, as well as some methods for solving them in one dimension. It also introduced the physical interpretation of these equations.\\
\end{adjustwidth}
Fetter, A. L., \& Walecka, J. D. (2003). \textit{Theoretical mechanics of particles and continua}. Mineola, NY: Dover Publications. \\
\begin{adjustwidth}{2 cm}{}
This book was used to understand the physical meaning behind the wave and heat equation, as well as some theorems from Sturm-Liouville theory.\\
\end{adjustwidth}
Jackson, J. D. (1972). Classical electrodynamics (1st ed.). New York: Wiley.\\
\begin{adjustwidth}{2 cm}{}
This book was useful in introducing the method of eigenfunction expansions, Green's functions, and different types of boundary value problems.\\
\end{adjustwidth}
Tolstov, G. P. (2009). \textit{Fourier series}. New York: Dover Publications, Inc.\\
\begin{adjustwidth}{2 cm}{}
This book was used to understand basic properties of Fourier Series and Fourier-Bessel Series, as well as general methods of eigenfunction expansions.
\end{adjustwidth}
\end{document}

>>>>

\documentclass{article}
\usepackage{changepage}
\usepackage{float}
\usepackage{noah}
\usepackage{{appendix}}

@@

    F_{net}=T \sin(\theta + d\theta)-T \sin(\theta)=T\cos(\theta)d\theta

@@

where $T$ is the tension in the rope. If we assume small enough displacements, we can approximate $\cos\theta\approx 1$ and write $d\theta= (\partial\theta/\partial x) dx$. Since we are assuming small displacements, we can say that the angle $\theta$ is the slope of the string. We will write the displacement of the string as $u=u(x,t)$, and say that this function must satisfy the following equation:

@@

Here we have absorbed the thermal conductivity into our definition of time. The interesting part of this equation is that it can be used to describe diffusive processes as well. In fact it can be used to describe many systems that do not violate the first and second law of thermodynamics. The first of these states energy conservation, and the second states that the entropy of a system must increase. This second law is one of the few that does depend on the direction of time. This is why the heat equation is not symmetric in the spatial and temporal variables, whereas the wave equation is.\\

@@

It is important to notice that for both of these equations, the spatial part is the same. Given that we are using the same boundary conditions for both equations, solving this equation will provide us with solutions to both the wave and heat equation.

@@

    \left(f(r,\phi),g(r,\phi)\right) = \int_0 ^1 dr \int_0 ^{2\pi} d\phi f^\star(r,\phi) g(r,\phi) r

@@

Before we continue, it is important to note that the normal modes that oscillate more spatially, (those with a larger eigenvalue) will dampen more quickly than others. This shows the smoothing property of solutions to the heat equation. Given that our boundary data states that the function must vanish at the boundary, we see that the steady state solution is one with uniform temperature. If the boundary was non-uniform, then we would find that the solution would evolve to a soution of Laplace's equation whose boundary values matched the boundary temperature.\\

@@

In a similar vein, we may write the solution to the wave equation in the following way:

@@

  \includegraphics[width = 1.0\textwidth]{images/wavevel.png}
  \caption{Evolution of wave equation for initial velocity given.}
We can see that these solutions to the heat equation are oscillating about the steady state solution $u = 0$. We also see that for the solution where only the initial velocity is given, the amplitude is much lower than it is when the initial displacement is given. This is because we are dividing each term of the solution by its corresponding eigenvalue. This in turn means that the initial velocity does not impact the dynamics as much as the initial position does.
\section{Conclusion}
Through the method of eigenfunction expansions, solutions to the heat and wave equation are generated, which in turn allow us to analyze the dynamics of solutions. We see that the wave equation is the higher dimensional analogue to the simple harmonic oscillator, and the heat equation is the higher dimensional analogue to exponential decay. Instead of the solutions oscillating or decaying towards an equilibrium point, they end up decaying to an equilibrium function, one that is harmonic.\\
It is important to reiterate that this method of solving functions by eigenfunction expansion is not unique to the Laplace operator, and can be used for any operator that satisfies equation (\ref{selfadjoint}), known as a self-adjoint operator. This allows us to in principal solve nearly any linear-homogenous PDE. In fact this method has a direct analogy to methods for solving a linear system of ODEs. Traditionally, we attempt to find the eigenvalues and straight-line solutions to the equation, which corresponds to finding the normal modes of a differential operator.
\begin{appendices}
\section{Solutions to Laplace's Equation}
Given their importance in both the heat and wave equations, it is important to introduce some properties of solutions to Laplace's equation. The most important property of Laplace's equation is the averaging property. Suppose we have a function $u:\Omega\rightarrow\bb{R}$, then for all $x\in\Omega$, $u(x)$ is the average of the outputs of $u$ for all points an equal distance from $x$. This is a feature we will prove for the unit disc.\\
In order to solve Laplace's equation, we begin with separation of variables:
  \frac{1}{f(r)}\left[ r^2 \dder{^2 f}{r^2} + r\dder{f}{r} \right] =-\frac{1}{g(\phi)}\dder{^2 g}{\phi^2} = n^2
Here the angular equation is the same as it was for the Helmholtz equation, however this time we express the solution in terms of sine and cosine functions
  g(\phi)=A_n \cos n\phi+B_n \sin n\phi
The radial equation is slightly different now since we no longer have the eigenvalue term:
  r^2 f''+ r f'- n^2 f=0
This is an Euler-Cauchy problem, and has solutions $r^{\pm n}$. Since we require the function to be at the very least continuous inside the unit disc, we only keep the $r^n$ term. This means that the general solution to Laplace's equation is:
  u(r,\phi)=\sum_{n=0}^\infty A_n r^n \cos n\phi + B_n r^n \sin n\phi
For the boundary conditions we have worked with through this report, we specified the value of the function on the boundary, so our boundary contition is $u(1,\phi)= h(\phi)$. If we insert this into our solution we get:
  h(\phi) = \sum_{n=0}^\infty A_n \cos n\phi+B_n \sin n\phi
This is the Fourier series in terms of sine and cosine functions. These form another basis for functions on the unit circle. If we rewrite the sine and cosine functions in terms of complex exponentials, we find the following relations between the expansion coefficients:
    A_o = 2 c_o\\
    A_n = c_n + c_{-n}\\
    B_n = i(c_n - c_{-n})
This allows us to write the expansion coefficients as:
    A_o = \frac{1}{2\pi}\int_0^{2\pi} u(1,\phi)d\phi\\
    A_n = \frac{1}{\pi}\int_0^{2\pi} u(1,\phi)\cos n\phi d\phi\\
    B_n = \frac{1}{\pi}\int_0^{2\pi} u(1,\phi)\sin n\phi d\phi\\
The solution $u(r,\phi)$ can be written in terms of a Green's function in the following way:
\begin{equation}P(r,\phi,\phi') = \frac{1}{2\pi}\left(1+ 2\sum_{n=1}^\infty r^n (\cos n\phi\cos n\phi'+\sin n\phi\sin n\phi') \right)
Through this function, we may write the solution as:
  u(r,\phi) = \int_0^{2\pi} u(1,\phi')P(r,\phi,\phi')d\phi'
Unlike the Green's functions for the wave or heat equations, this can be simplified greatly. We first make use of the identity $\cos(a-b)=\cos a\cos b+\sin a\sin b$, and then rewrite our cosine function in terms of complex exponentials,
  P(r,\phi,\phi') = \frac{1}{2\pi}\left(
    1+\sum_{n = 1}^{\infty } (re^{i(\phi-\phi')})^n + (re^{-i(\phi-\phi')})^n 
  \right)
The geometric series tells us that:
  \sum_{n=0}^\infty x^n = 1+\sum_{n=1}^\infty x^n = \frac{1}{1-x}
for $|x|<1$. Since $r\in [0,1]$ and the complex exponentials never exceed 1, we have convergence:
  P(r,\phi,\phi') = \frac{1}{2\pi}\left(1+\frac{1}{1-r \exp(i(\phi-\phi'))}-1 + \frac{1}{1-\exp(-i(\phi-\phi'))} - 1\right)
This can be simplified to give:
    P(r,\phi,\phi') = \frac{1}{2\pi} \frac{1-r^2}{1 - 2r \cos(\phi-\phi') - r^2}
This is known as the Poisson kernel, and allows us to write a general solution to Laplace's equation. If we consider the value of the function at $r=0$, we see that the value is:
  u(0,\phi) = \frac{1}{2\pi}\int_0^{2\pi} u(1,\phi)d\phi
which confirms the averaging property of solutions to Laplace's equation.
\section{Inhomogenous Boundary Conditions}
Throughout this report, we used the boundary condition that the function must vanish at the boundary. However in certain situations, the boundary may be a non-zero function for all time. In the contex of the the heat equation, this means that the boundary is kept at a fixed temperature, and the solution will tend towards the solution to Laplace's equation with that boundary value. Suppose that for both the wave and heat equation, we have the restriction $u(1,\phi,t)=h(\phi)$. We will define a new function:
  v(r,\phi,t) = u(r,\phi,t) - \int_0^{2\pi} P(r,\phi,\phi')h(\phi')d\phi'
where $P$ is the Poisson kernel. We see that $v$ is now zero at the boundary, and still will satisfy the wave or heat equation. We may now solve the initial value problem for $v(r,\phi,t)$ and then solve for $u(r,\phi,t)$. 
\end{appendices}
\section{References}
Asmar, N. H. (2002). \textit{Applied complex analysis with partial differential equations}. New Jersey: Prentice-Hall Inc.\\
\begin{adjustwidth}{2 cm}{}
This book was used to understand differential equations in polar and spherical coordinate systems, as well as certain properties of Bessel functions.\\
\end{adjustwidth}
Farlow, S. J. (2016). \textit{Partial differential equations for scientists and engineers}. New York: Dover Publications, Inc.\\
\begin{adjustwidth}{2 cm}{}
This book was used to understand the basic types of partial differential equations, as well as some methods for solving them in one dimension. It also introduced the physical interpretation of these equations.\\
\end{adjustwidth}
Fetter, A. L., \& Walecka, J. D. (2003). \textit{Theoretical mechanics of particles and continua}. Mineola, NY: Dover Publications. \\
\begin{adjustwidth}{2 cm}{}
This book was used to understand the physical meaning behind the wave and heat equation, as well as some theorems from Sturm-Liouville theory.\\
\end{adjustwidth}
Jackson, J. D. (1972). Classical electrodynamics (1st ed.). New York: Wiley.\\
\begin{adjustwidth}{2 cm}{}
This book was useful in introducing the method of eigenfunction expansions, Green's functions, and different types of boundary value problems.\\
\end{adjustwidth}
Tolstov, G. P. (2009). \textit{Fourier series}. New York: Dover Publications, Inc.\\
\begin{adjustwidth}{2 cm}{}
This book was used to understand basic properties of Fourier Series and Fourier-Bessel Series, as well as general methods of eigenfunction expansions.
\end{adjustwidth}
\end{document}
